Supplemental Information

### 2025-03-13 paper/paper.Rmd
si cleanup

We retrieved data from Peekbank, focusing on data from English-speaking children ages 1--6  years and on simple word recognition trials in which children were shown two pictures of concrete objects and heard a label for an object (typically embedded in a simple carrier phrase such as "Look at the ..."). These criteria yielded `r length(unique(d_sub$dataset_name))` datasets, including `r length(unique(d_sub$subject_id))` children and `r length(unique(d_sub$administration_id))` administrations of the LWL procedure (some datasets were longitudinal or involved multiple closely-spaced testing sessions). Table \ref{tab:datasets} shows the characteristics of individual datasets (see also Figures S1 and S2). The size of the combined dataset, the unified data processing pipeline, and the fact that individual studies used very similar implementations of the LWL experimental paradigm, all allowed us to make a more detailed study of the development of word recognition than has previously been possible. While our analysis is exploratory in nature, it is guided by the two hypotheses outlined above: the presence of 1) signatures of skill learning in word recognition, and 2) linkages between word recognition and vocabulary.

@@

We began by characterizing age gradients in speed and accuracy across children. We computed both RTs (reaction times) and accuracies (proportion looking at the target image) following standard practices in the literature [@fernald2008]. Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2000 ms after noun onset) and a longer window (from 200 -- 4000 ms). For each window, we averaged all fixation within the window to compute a continuous proportion target looking between 0 (no fixation on the target during the window) and 1 (total fixation on the target during the window) on every trial. In this initial analysis, we treat observations of RT and target looking as direct measures of the constructs speed and accuracy; in subsequent analyses we estimate latent variables representing these constructs.

Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI for raw correlations between variables). We began by fitting linear mixed effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age by study and random intercepts by participant. We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT. The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age,^[This longer window yielded overall higher cross-trial reliability as well, and so we report long window results in the main text. All findings reported here hold for both window sizes, however (see SI).] and the best fitting model of speed predicted log RT as a function of log age as well (see SI).

@@

On the other hand, it is possible that differences in predicted growth trajectories are due to overall coupling between vocabulary size and language processing, rather than a specifically predictive relationship between $t_0$ RT and vocabulary growth. To test this relationship, we used longitudinal structural equation models. We separated the longitudinal speed, accuracy, and vocabulary data into two-month bins spanning up to 10 months ($t_0, ..., t_4$) and fit individual growth across each of these variables. We used full-information maximum likelihood to handle the substantial missing data caused by the different longitudinal sampling schemes of studies in our dataset. (We present a model of coupled growth in the observed variables here for simplicity; see SI for a comparable model using growth in the latent factors). The fitted longitudinal model is shown in Figure \ref{fig:longitudinal}. Overall fit statistics were generally acceptable (Confirmatory fit index: `r round(slopes_long_summary$fit["cfi"], 3)`, RMSE: `r round(slopes_long_summary$fit["rmsea"], 3)`, RMSE $p$-value: `r round(slopes_long_summary$fit["rmsea.pvalue"], 3)`).

>>>>

We retrieved data from Peekbank, focusing on data from English-speaking children ages 1--6  years and on simple word recognition trials in which children were shown two pictures of concrete objects and heard a label for an object (typically embedded in a simple carrier phrase such as "Look at the ..."). These criteria yielded `r length(unique(d_sub$dataset_name))` datasets, including `r length(unique(d_sub$subject_id))` children and `r length(unique(d_sub$administration_id))` administrations of the LWL procedure (some datasets were longitudinal or involved multiple closely-spaced testing sessions). Table \ref{tab:datasets} shows the characteristics of individual datasets (see also SI: Dataset Description). The size of the combined dataset, the unified data processing pipeline, and the fact that individual studies used very similar implementations of the LWL experimental paradigm, all allowed us to make a more detailed study of the development of word recognition than has previously been possible. While our analysis is exploratory in nature, it is guided by the two hypotheses outlined above: the presence of 1) signatures of skill learning in word recognition, and 2) linkages between word recognition and vocabulary.

@@

We began by characterizing age gradients in speed and accuracy across children. We computed both RTs (reaction times) and accuracies (proportion looking at the target image) following standard practices in the literature [@fernald2008]. Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2000 ms after noun onset) and a longer window (from 200 -- 4000 ms). For each window, we averaged all fixation within the window to compute a continuous proportion target looking between 0 (no fixation on the target during the window) and 1 (total fixation on the target during the window) on every trial. In this initial analysis, we treat observations of RT and target looking as direct measures of the constructs speed and accuracy (see SI: Test-Retest Reliability); in subsequent analyses we estimate latent variables representing these constructs.

Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI: Correlations for raw correlations between variables). We began by fitting linear mixed effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age by study and random intercepts by participant. We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT. The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age,^[This longer window yielded overall higher cross-trial reliability as well, and so we report long window results in the main text. All findings reported here hold for both window sizes, however (see SI: Short Window Analysis).] and the best fitting model of speed predicted log RT as a function of log age as well (see SI: Functional Form Model Comparison).

@@

On the other hand, it is possible that differences in predicted growth trajectories are due to overall coupling between vocabulary size and language processing, rather than a specifically predictive relationship between $t_0$ RT and vocabulary growth. To test this relationship, we used longitudinal structural equation models. We separated the longitudinal speed, accuracy, and vocabulary data into two-month bins spanning up to 10 months ($t_0, ..., t_4$) and fit individual growth across each of these variables. We used full-information maximum likelihood to handle the substantial missing data caused by the different longitudinal sampling schemes of studies in our dataset. (We present a model of coupled growth in the observed variables here for simplicity; see SI: Latent Factor Growth for a comparable model using growth in the latent factors). The fitted longitudinal model is shown in Figure \ref{fig:longitudinal}. Overall fit statistics were generally acceptable (Confirmatory fit index: `r round(slopes_long_summary$fit["cfi"], 3)`, RMSE: `r round(slopes_long_summary$fit["rmsea"], 3)`, RMSE $p$-value: `r round(slopes_long_summary$fit["rmsea.pvalue"], 3)`).

### 2025-03-13 paper/supplemental_information.Rmd
lots of renaming
si cleanup

# Dataset description

@@

ggplot(d_sub_long, 
       aes(x = delta_t, fill = dataset_name)) +
  geom_histogram(binwidth = 1) + 
  # facet_wrap(~dataset_name) + 
  xlab("Age since first test")
```
# Test retest reliability

@@

# Speed and accuracy

Correlations between variables

@@

@@

# Exploratory Factor Analysis

- Full EFA results

@@

# Sensitivity Analysis for Longitudinal Models

THIS IS WRONG

```{r reliability_table, results="asis"}
xtable(dataset_reliabilities)
```

Sensitivity analysis for age split point (e.g., split at 3 months)

Sensitivity analysis for measurement invariance (e.g., not constraining loadings)

# Analyses using short window accuracy

## Longitudinal factor analysis

## Longitudinal factor analysis

>>>>

# Dataset Description

@@

ggplot(d_sub_long, 
       aes(x = delta_t, fill = dataset_name)) +
  geom_histogram(binwidth = 1) + 
  # facet_wrap(~dataset_name) + 
  xlab("Age since first test")
```
# Test-Retest Reliability

@@

# Correlations

@@

# Functional Form Model Comparison

@@

With our current dataset we cannot specifically determine whether within-individual patterns of change conform to power law or exponential forms [@kail1991,@heathcote2000], but regardless we see clear patterns of developmental change related to skill acquisition.

# Factor Analysis

```{r}
d_sub_mat <- d_sub |>
  ungroup() |>
  select(dataset_name, rt, rt_var, long_window_accuracy, long_window_acc_var, prod, comp, age)

d_sub_mat_s <- d_sub_mat |>
  ungroup() |>
  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy", 
                         "long_window_acc_var", "prod", "comp")), 
                       ~ age_scale(.x, age))) 
fa.parallel(select(d_sub_mat, -dataset_name, -age), fa = "fa", 
            use = "pairwise.complete.obs")
```

```{r}
fs <- fa(select(d_sub_mat, -dataset_name), nfactor = 3,
            use = "pairwise", rotate = "varimax")

as.data.frame(unclass(loadings(fs))) |> 
  kable(caption = "Factor loadings for the exploratory three factor solution using varimax rotation.", digits = 2)
```

@@

# Latent Factor Growth

```{r}
fa3_model_long <- "
# measurement model
vocab_t1 =~ 1*prod_t1 + s1*comp_t1
accuracy_t1 =~ 1*acc_t1 + s2*acc_sd_t1
speed_t1 =~ 1*log_rt_t1 + s3*log_rt_sd_t1

vocab_t2 =~ 1*prod_t2 
accuracy_t2 =~ 1*acc_t2 + s2*acc_sd_t2
speed_t2 =~ 1*log_rt_t2 + s3*log_rt_sd_t2

vocab_t3 =~ 1*prod_t3 
accuracy_t3 =~ 1*acc_t3 + s2*acc_sd_t3
speed_t3 =~ 1*log_rt_t3 + s3*log_rt_sd_t3

vocab_t4 =~ 1*prod_t4 
accuracy_t4 =~ 1*acc_t4 + s2*acc_sd_t4
speed_t4 =~ 1*log_rt_t4 + s3*log_rt_sd_t4

# means for the latents
# vocab_t1 ~ 1
# vocab_t2 ~ 1
# vocab_t3 ~ 1
# vocab_t4 ~ 1
# accuracy_t1 ~ 1
# accuracy_t2 ~ 1
# accuracy_t3 ~ 1
# accuracy_t4 ~ 1
# speed_t1 ~ 1
# speed_t2 ~ 1
# speed_t3 ~ 1
# speed_t4 ~ 1

# autoregression for the latents
# vocab_t2 ~ 1 * vocab_t1
# vocab_t3 ~ 1 * vocab_t2
# vocab_t4 ~ 1 * vocab_t3
# accuracy_t2 ~ 1 * accuracy_t1
# accuracy_t3 ~ 1 * accuracy_t2
# accuracy_t4 ~ 1 * accuracy_t3
# speed_t2 ~ 1 * speed_t1
# speed_t3 ~ 1 * speed_t2
# speed_t4 ~ 1 * speed_t3

# regressions
accuracy_intercept =~ 1*accuracy_t1 + 1*accuracy_t2 + 1*accuracy_t3 + 1*accuracy_t4 
accuracy_slope =~ 1*accuracy_t1 + 2*accuracy_t2 + 3*accuracy_t3 + 4*accuracy_t4
speed_intercept =~ 1*speed_t1 + 1*speed_t2 + 1*speed_t3 + 1*speed_t4 
speed_slope =~ 1*speed_t1 + 2*speed_t2 + 3*speed_t3 + 4*speed_t4
vocab_intercept =~ 1*vocab_t1 + 1*vocab_t2 + 1*vocab_t3 + 1*vocab_t4 
vocab_slope =~ 1*vocab_t1 + 2*vocab_t2 + 3*vocab_t3 + 4*vocab_t4

# parameters for the regressions
vocab_intercept ~ 1
vocab_slope ~ 1
accuracy_intercept ~ 1
accuracy_slope ~ 1
speed_intercept ~ 1
speed_slope ~ 1

# residual variance for the latents
vocab_t1 ~~ NA*vocab_t1
vocab_t2 ~~ NA*vocab_t2
vocab_t3 ~~ NA*vocab_t3
vocab_t4 ~~ NA*vocab_t4
accuracy_t1 ~~ NA*accuracy_t1
accuracy_t2 ~~ NA*accuracy_t2
accuracy_t3 ~~ NA*accuracy_t3
accuracy_t4 ~~ NA*accuracy_t4
speed_t1 ~~ NA*speed_t1
speed_t2 ~~ NA*speed_t2
speed_t3 ~~ NA*speed_t3
speed_t4 ~~ NA*speed_t4

# residual variance for the other latents
vocab_intercept ~~ NA*vocab_intercept
vocab_slope ~~ NA*vocab_slope
accuracy_intercept ~~ NA*accuracy_intercept
accuracy_slope ~~ NA*accuracy_slope
speed_intercept ~~ NA*speed_intercept
speed_slope ~~ NA*speed_slope

# set observed intercepts to zero - these should go into the latents
prod_t1 ~ 0
prod_t2 ~ 0
prod_t3 ~ 0
prod_t4 ~ 0
comp_t1 ~ 0 
log_rt_t1 ~ 0
log_rt_t2 ~ 0
log_rt_t3 ~ 0
log_rt_t4 ~ 0
log_rt_sd_t1 ~ 0 
log_rt_sd_t2 ~ 0
log_rt_sd_t3 ~ 0 
log_rt_sd_t4 ~ 0 
acc_t1 ~ 0 
acc_t2 ~ 0 
acc_t3 ~ 0 
acc_t4 ~ 0 
acc_sd_t1 ~ 0
acc_sd_t2 ~ 0
acc_sd_t3 ~ 0
acc_sd_t4 ~ 0
"

fit3_long <- sem(fa3_model_long, d_sub_wide, std.lv=TRUE, missing='fiml')

summary(fit3_long, fit.measures=TRUE, standardize=TRUE)

```{r}
layout <- read.csv(here("misc/layout_dt3.csv"), header = FALSE)

graph_sem(model = fit3_long, text_size = 3, layout=t(layout))

# Short Window Analysis

Need to redo all figures using short windows ugh.
